Measurement of total hadronic differential cross sections in the LArIAT experiment

### 2018-03-20 Chapter-2/2-LArTPCs.tex
s

\subsection{Time Projection Chamber}
\subsection{Ionization Detectors with Noble Liquids}
\subsection{LArTPC: Principles of Operation}
\subsection{Liquid Argon Ionization Charge Detection}
\subsubsection{Electron Life Time \& purity}
\subsubsection{Space Charge Effect}
\subsubsection{Recombination Effect}
\subsection{Liquid Argon scintillation Light Detection}
\subsubsection{LAr Scintillation Process}
\subsubsection{Wavelength Shifting of LAr Scintillation Light}
\section{The SBN Program: Neutrino Interaction and Detection}
%\subsection{SBN Goals}
%\subsection{Neutrino Interactions and Detection }
\section{DUNE: Rare Decay Searches}
%\subsection{Non-Accelerator Physics Program}
%\subsection{Rare Decay Searches: Experimental Limit}
%\subsection{Nucleon Decay Detection in LAr}
\section{Enabling the next generation of discoveries: LArIAT}

>>>>

\subsection{Time Projection Chamber}
\subsection{Ionization Detectors with Noble Liquids}
\subsection{LArTPC: Principles of Operation}
\subsection{Liquid Argon Ionization Charge Detection}
\subsubsection{Electron Life Time \& purity}
\subsubsection{Space Charge Effect}
\subsubsection{Recombination Effect}
\subsection{Liquid Argon scintillation Light Detection}
\subsubsection{LAr Scintillation Process}
\subsubsection{Wavelength Shifting of LAr Scintillation Light}
\section{The SBN Program: Neutrino Interaction and Detection}
%\subsection{SBN Goals}
%\subsection{Neutrino Interactions and Detection }
\section{DUNE: Rare Decay Searches}
%\subsection{Non-Accelerator Physics Program}
%\subsection{Rare Decay Searches: Experimental Limit}
%\subsection{Nucleon Decay Detection in LAr}
\section{Enabling the next generation of discoveries: LArIAT}
LArIAT, a small Liquid Argon Time Projection Chamber (LArTPC) in a test beam,  is designed to perform an extensive physics campaign centered on charged particle cross section measurements while characterizing the detector performance for future LArTPCs. LArTPC represents one of the most advanced experimental technologies for physics at the Intensity Frontier due to its full 3D-imaging, excellent particle identification and precise calorimetric energy reconstruction. This complex technology however needs a thorough calibration and dedicated measurements of some key quantities to achieve the precision required for the next generation of discoveries at the Intensity Frontier which LArIAT can provide.

The LArIAT LArTPC is deployed in a dedicated calibration test beamline at Fermilab.
We use the LArIAT beamline to characterize the charge particles before they enter the TPC: the particle type and initial momentum is known from beamline information. The precise calorimetric energy reconstruction of the LArTPC technology enables the measurement of the total differential cross section for  tagged hadrons. 
The Pion-Nucleus and Kaon-Nucleus total hadronic interaction cross section have never been measured before in argon and they are a fundamental step to shed light on light meson interaction in nuclei. Additionally, these measures provides a key input to neutrino physics and proton decay studies in future LArTPC experiments like SBN and DUNE.
\textcolor{red}{add paragraph on all wonderful things lariat can do... some event displays would be nice!}

### 2018-03-20 Chapter-3/3-TheLArIATExperiment.tex
s

\chapter{LArIAT: Liquid Argon In A Testbeam}\label{sec:experimentDescription}

LArIAT, a small Liquid Argon Time Projection Chamber (LArTPC) in a test beam,  is designed to perform an extensive physics campaign centered on charged particle cross section measurements while characterizing the detector performance for future LArTPCs. LArTPC represents one of the most advanced experimental technologies for physics at the Intensity Frontier due to its full 3D-imaging, excellent particle identification and precise calorimetric energy reconstruction. This complex technology however needs a thorough calibration and dedicated measurements of some key quantities to achieve the precision required for the next generation of discoveries at the Intensity Frontier which LArIAT can provide.

The LArIAT LArTPC is deployed in a dedicated calibration test beamline at Fermilab.
We use the LArIAT beamline to characterize the charge particles before they enter the TPC: the particle type and initial momentum is known from beamline information. The precise calorimetric energy reconstruction of the LArTPC technology enables the measurement of the total differential cross section for  tagged hadrons. 
The Pion-Nucleus and Kaon-Nucleus total hadronic interaction cross section have never been measured before in argon and they are a fundamental step to shed light on light meson interaction in nuclei. Additionally, these measures provides a key input to neutrino physics and proton decay studies in future LArTPC experiments like SBN and DUNE.

@@

The Fermilab accelerator complex works in supercycles of roughly 60 seconds in duration. The beam is then split by electrostatic septa and delivered at different experimental halls all over the lab. A 120~GeV$/c$ primary proton beam with variable intensity is extracted in four-second ``spills" and sent to the Meson Center beam line. Here, this primary beam is focused onto a tungsten target to create LArIAT's secondary beam. The composition of the secondary particle beam is mainly positive pions. For the LArIAT data considered in this work the secondary beam peak momentum was fixed at 64~GeV$/c$, although the beam is tunable in momentum between 8-80\,GeV$/c$; this was deemed a secondary beamline configuration which allowed a stable beam operation of the FTBF.
The beam of pions impinges then on a copper target within a steel collimator inside the LArIAT experimental hall (MC7) to create LArIAT tertiary beam, whose geometry in MC7 has been optimized for LArIAT (shown in  Fig.~\ref{fig:tert-layout}).   The steel collimator selects particles produced with a $13^\circ$ production angle at the target down the beamline.  The particles are then bend by  $~10^\circ$  through a pair of dipole magnets.  Tuning the magnets field intensity results in a range of particle momenta from 0.2 to 1.4~GeV/c.
The tertiary beam composition counts mostly pions and protons with a small fraction of electrons, muons, and kaons present as well. It is the job of the LArIAT beamline detectors to select the particles polarity,  to perform particle identification (beamPID) and to measure the momentum of the tertiary beam particles before they get to the LArTPC. The LArIAT detectors are described in the following paragraphs.

@@

Two ASDQ chips~\cite{ASDQchip} mounted on a mother board plugged into the chamber serve as front end amplifier/discriminator. The chips are connected to a multi-hit TDC~\cite{Sten} which provides a fast OR output used as first level trigger. The TDC time resolution is 1.18~ns/bin and can accept 2 edges per 9~ns.  
The maximum event rate acceptable by the chamber system is of 1 MHz: this rate is not a limiting factor considering that the rate of the tertiary particle beam at the first wire chamber is estimated to be less than 15 kHz. A full spill of data occurring once per supercycle is stored on the TDC board memory at once and read out by a controller specially designed controller.  We use LVDS cables to carry both power and data between the controller and the TDCs and from the controller to the rest of the DAQ.  
%It is possible to program the time window for acceptance for hits, time offsets, front end threshold, and pulse shaping parameters through the controller via a USB from a PC or through an Ethernet connection.

\subsubsection{Multi-Wire Proportional Chambers functionality}\label{sec:MWPCfunc}
We use the wire chamber system together with the bending magnets to measure the particles momentum.

In the simplest scenario, only one hit on each and every of the four wire chambers is recorder during a single readout of the detector systems.  Thus, we use the hits position in the two wire chambers upstream to the magnets to form a trajectory before the bent and the hits position in the two wire chambers downstream to the magnets to form a trajectory after the bent. We use the angles in the XZ plane between the upstream and downstream trajectories  to calculate the z component of the momentum as follows:

@@

where $B_{eff}$ is the effective maximum field in a square field approximation,  $L_{eff}$ is the effective length of both magnets (twice the effective length of one magnet) and  3.3~$c^{-1}$ the conversion factor from from T*m to MeV/c. By using the hit positions on WC3 and WC4, we estimate the azimutal and polar angles of the particle trajectory, and we are able to calculate the other components of the momentum.

This simple scenario can get complicated in two ways:  the presence of multiple hits in a single wire chamber or by the absence of hits in one (or more) wire chamber. The first complication is due to beam pile up, while the latter is due to wire chamber inefficiency. Even in the case of multiple hits, at most one wire chamber track is reconstructed per event; since the magnets bend particles only in the X direction, we assume the particle trajectory to be roughly constant in the YZ plane, thus we keep the combination of hits most coplanar on YZ plane.  It is possible to reconstruct the particle momentum even in case of wire chamber inefficiency, if the information is missing in either of the two middle wire chambers (WC2 or WC3). %Assuming identical magnets, the upstream and downstream trajectories projected without bending should intersect at some point in the space between the magnets. Over a collection of many tracks, these points of intersection would trace a plane, therefore the term "midplane" is used. The completed half of the wire chamber track is projected to the midplane and the point of intersection is used with the incomplete half to finish the track.

\textcolor{red}{BETTER EXPLAINATION OF HIGH YIELD}
\textcolor{red}{UNCERTAINTY ON MOMENTUM MEASUREMENT}
Events satisfying the simplest scenario of one single hit in each of the four wire chambers form the so called ``Picky track" sample.  We construct an other, higher statistics sample where we loose the requirements on single hit and wire chamber efficiency called "High Yield". For LArIAT Run II, the "High Yield" sample is 
\textcolor{red}{ HIGH YIELD increase in stat} the "Picky Tracks" sample. We use Picky tracks for calibration purposes: \textcolor{red}{ I DONT UNDERSTAND THIS Plotting the fractional error of the corrected three-point momentum compared to the known four-point momentum shows an uncertainty of approximately 2\%.}

@@

\section{Trigger and DAQ}
The LArIAT DAQ and trigger system governs the read out of all the many detectors forming LArIAT. 
The CAEN V1495 module featuring a user-programmable FPGA  is the core of this system.  Every 10~ns, this module checks for matches between sixteen logical inputs and user-defined patterns in the trigger menu; if it finds a match for two consecutive clock ticks, that trigger fires.

% converts signals from analog to digital, interfaces and synchronizes with the accelerator system, stores raw data on disk, transfers data to archival storage and conduct a first pass data quality monitoring.

%A CAEN V1495 board is used to perform the trigger of the whole detectors system. It is a VME board, which can be directly customized by the user through two programmable FPGA. For LArIAT experiment 16 trigger input signals are used, see Fig. 2.14. Through the FPGA configuration file  V1495.config  it is possible to define different trigger conditions - trigger output paths, with several logic combinations of the input signals.
%The former trigger output path is the one that selects particles from the beam that cross the whole beamline and get to the TPC.
%Moreover it's possible to add a delay to the trigger and to set some combinations of signals as a VETO.
%Waveforms from beamline detectors and optical devices (PMTs and SiPMs) inside the cryostat are collected using two CAEN V1751 boards (For v1751 boards inputs, see Table 2.2.). They are both fed by the V1495 fast trigger. Each has a unit boardId and produces CAENFragment raw data objects.
%The CAEN V1751 Waveform Digitizer [40] is VME module housing 8 Channel 10 bit 1 GS/s Waveform Digitizer with 1 Vpp input dynamic range; the range that has been set for our purposes is 204 mV, so 1 ADC count corresponds to 0.2 mV. The input channel provides a programmable voltage o set in the ± 0.5 V range too; we used this feature to adjust baseline o sets on our signals. The time window can also be adjusted by user maintaining the same sample %frequency (1 GS/s); we've started with a 7.168 ?s
%time window, but after preliminary studies on the signals collected by PMTs and SiPMs we've moved to a 14.336 ?s window that could allow us to sample over all the slow LAr scintillation component (?S) signal tails, to study single photoelectron response of the optical devices.

%The LArIAT data acquisition system is triggered to read out the digitizing buffers by a CAEN V1495 module. The V1495 is a powerful, easily configurable coincidence module featuring a user-programmable FPGA.  Sixteen logical inputs  are sampled on a 10~ns clock, checking for matches to any of the sixteen user-defined patterns in the trigger menu.  If a given trigger pattern is satisfied for two consecutive clock ticks, that pattern fires.

%NIM-standard logic pulse inputs to the trigger decision come from any of the instruments in the experiment, in principle; trigger inputs are derived from beam instruments, from the cosmic ray taggers, and from the cryostat's scintillation detectors. LArIAT takes full advantage of the trigger card's great flexibility.  As conditions change and innovations are gradually made to LArIAT's data-taking process, improvements are made to the trigger input designs and the combinations of inputs which define trigger patterns for the V1495.  The configuration of the trigger inputs is automatically logged in a database at the start of each run as part of the DAQ configuration, along with the rest of the XML configuration file for the V1495.

%\subsubsection{Inputs to the Trigger System}

@@

>>>>

\chapter{LArIAT: Liquid Argon In A Testbeam}\label{sec:experimentDescription}
In this chapter, we describe the LArIAT experimental setup. We start by illustrating the journey of the charge particles in the Fermilab accelerator complex, from gaseous thermal hydrogen in \textcolor{red}{find place where hydrogen is} to the delivery of the LArIAT tertiary beam to our experimental hall. We  then describe the LArIAT beamline detectors, the LArTPC, the DAQ and the monitoring system.

@@

The Fermilab accelerator complex works in supercycles of roughly 60 seconds in duration. The beam is then split by electrostatic septa and delivered at different experimental halls all over the lab. A 120~GeV$/c$ primary proton beam with variable intensity is extracted in four-second ``spills" and sent to the Meson Center beam line. Here, this primary beam is focused onto a tungsten target to create LArIAT's secondary beam. The composition of the secondary particle beam is mainly positive pions. For the LArIAT data considered in this work the secondary beam peak momentum was fixed at 64~GeV$/c$, although the beam is tunable in momentum between 8-80\,GeV$/c$; this was deemed a secondary beamline configuration which allowed a stable beam operation of the FTBF.
The beam of pions impinges then on a copper target within a steel collimator inside the LArIAT experimental hall (MC7) to create LArIAT tertiary beam, whose geometry in MC7 has been optimized for LArIAT (shown in  Fig.~\ref{fig:tert-layout}).   The steel collimator selects particles produced with a $13^\circ$ production angle at the target down the beamline.  The particles are then bend by  $~10^\circ$  through a pair of dipole magnets.  Tuning the magnets field intensity results in a range of particle's momenta from 0.2 to 1.4~GeV/c.
The tertiary beam composition counts mostly pions and protons with a small fraction of electrons, muons, and kaons present as well. It is the job of the LArIAT beamline detectors to select the particles polarity,  to perform particle identification (beamPID) and to measure the momentum of the tertiary beam particles before they get to the LArTPC. The LArIAT detectors are described in the following paragraphs.

@@

Two ASDQ chips~\cite{ASDQchip} mounted on a mother board plugged into the chamber serve as front end amplifier/discriminator. The chips are connected to a multi-hit TDC~\cite{Sten} which provides a fast OR output used as first level trigger. The TDC time resolution is 1.18~ns/bin and can accept 2 edges per 9~ns.  
The maximum event rate acceptable by the chamber system is of 1 MHz: this rate is not a limiting factor considering that \textcolor{red}{the rate of the tertiary particle beam at the first wire chamber is estimated to be less than 15 kHz}. A full spill of data occurring once per supercycle is stored on the TDC board memory at once and read out by a specially designed controller.  We use LVDS cables to carry both power and data between the controller and the TDCs and from the controller to the rest of the DAQ.  
%It is possible to program the time window for acceptance for hits, time offsets, front end threshold, and pulse shaping parameters through the controller via a USB from a PC or through an Ethernet connection.

\subsubsection{Multi-Wire Proportional Chambers functionality}\label{sec:MWPCfunc}
We use the wire chamber system together with the bending magnets to measure the particle's momentum.

In the simplest scenario, only one hit on each and every of the four wire chambers is recorded during a single readout of the detector systems.  Thus, we use the hit positions in the two wire chambers upstream of the magnets to form a trajectory before the bend, and the hit positions in the two wire chambers downstream of the magnets to form a trajectory after the bend. We use the angles in the XZ plane between the upstream and downstream trajectories  to calculate the z component of the momentum as follows:

@@

where $B_{eff}$ is the effective maximum field in a square field approximation,  $L_{eff}$ is the effective length of both magnets (twice the effective length of one magnet), $\theta_{US}$ is the angle off the $z$ axis of the upstream trajectory, $\theta_{DS}$ is the angle off the $z$ axis of the downstream trajectory  and  3.3~$c^{-1}$ is the conversion factor from [T$\cdot$m] to [MeV/c]. By using the hit positions on WC3 and WC4, we estimate the azimutal and polar angles of the particle trajectory, and we are able to calculate the other components of the momentum.

This simple scenario can get complicated in two ways:  the presence of multiple hits in a single wire chamber or by the absence of hits in one (or more) wire chambers. The first complication is due to beam pile up, while the latter is due to wire chamber inefficiency. Even in the case of multiple hits, at most one wire chamber track is reconstructed per event. Since the magnets bend particles only in the X direction, we assume the particle trajectory to be roughly constant in the YZ plane, thus we keep the combination of hits which fit best with a straight line.  Even if the information is missing in either of the two middle wire chambers (WC2 or WC3), it is still possible to reconstruct the particle's momentum. Under the assumption of identical magnets, we define a plane centered in the middle of the two magnets (called ``midplane") that the physical particles need to cross. We project the completed half of the wire chamber track to the midplane, assuming no bending, to find the point of intersection. We use this point to complete the other half of the wire chamber track and to calculate the reconstructed momentum  with these four points. To account for the lack of bending in our reconstructed wire chamber track, we apply to the calculated momentum  a correction obtained with a sample of 4-point, single hit tracks.

Events satisfying the simplest scenario of one single hit in each of the four wire chambers form the ``Picky track" sample.  We construct another, higher statistics sample, where we loosen the requirements on single hit and wire chamber efficiency: the ``High Yield" sample. For LArIAT Run II, the High Yield sample is about three times the Picky Tracks statistics. We use Picky Tracks for calibration purposes, in particular to obtain the momentum correction for the 3-point tracks; this correction adds an uncertainty of approximately 2\% to the momentum calculation of the High Yield sample.

\textcolor{red}{Four point track momentum uncertainty}

@@

\section{Trigger and DAQ}
The LArIAT DAQ and trigger system governs the read out of all the many detectors forming LArIAT. 
The CAEN V1495 module and its user-programmable FPGA  are the core of this system.  Every 10~ns, this module checks for matches between sixteen logical inputs and user-defined patterns in the trigger menu; if it finds a match for two consecutive clock ticks, that trigger fires.

The beam instruments,  the cosmic ray taggers, and the light collection system provide NIM-standard logic pulse inputs to the trigger decision. We automatically log the trigger inputs configuration with the rest of the DAQ configuration at the beginning of each run.

%%%%%%%%%%%%%%%%%%%%%%%%%

%
%\subsubsection{Inputs to the Trigger System}

@@

%%%%%%%%%%%%%%%%%%%%%%
